Weighted Model Counting Without Parameter Variables

### 2021-03-08 doc/paper3/paper.tex
First draft ready

\begin{abstract} % TODO: write
  The abstract should briefly summarize the contents of the paper in
  150--250 words.

  \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

\section{Introduction}
% TODO for later:
% 1) read through v2 of the paper and check if I forgot to mention something
% here
% 2) run this through grammarly

@@

% NOTE: What's wrong with previous proposed solutions? How does mine differ?
One approach to this problem considers weighted clauses and probabilistic
semantics based on Markov networks \cite{DBLP:conf/uai/GogateD10}. However, with
a new representation comes the need to invent new encodings and inference
algorithms. Our work is similar in that it introduces a new representation for
computational problems but is able to reuse recent WMC algorithms based on
pseudo-Boolean function manipulation, namely, \textsf{ADDMC}
\cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC} \cite{DBLP:conf/cp/DudekPV20}.
Furthermore, we identify necessary conditions for transforming a WMC instance
into the new format. As many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}, another way to avoid parameter variables
could be via direct compilation to a more convenient representation. However,
while direct compilation of Bayesian networks to SDDs has been investigated
\cite{DBLP:conf/ecsqaru/ChoiKD13}, SDDs are not expressive enough to avoid the
issue.

% NOTE: What are the key components of my approach and results? Also include any
% specific limitations. What about your project is new or different from
% existing solutions? List of refutable contributions (with forward references
% to the rest of the paper) What is the outcome of the project?
In this work, we introduce a way to transform WMC problems to a new format based
on pseudo-Boolean functions---\emph{pseudo-Boolean projection} (PBP). We
formally show that every WMC problem instance has a corresponding PBP instance
and identify conditions under which this transformation can remove parameter
variables. Four out of the five known WMC encodings for Bayesian networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}
can indeed be simplified in this manner. This transformation enables two
encodings that were previously incompatible with most WMC algorithms (due to
using a different definition of WMC
\cite{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}) to be run with
\textsf{ADDMC} and \textsf{DPMC} and results in a significant performance boost
for one other encoding, allowing it to surpass the current state of the art.
Finally, our theoretical contributions result in a convenient algebraic way of
reasoning about two-valued pseudo-Boolean functions and position WMC encodings
on a common ground, identifying their key properties and assumptions.

%To the best of the authors' knowledge, none of these compilation/inference
%techniques have a publicly available implementation.

@@

We begin with an overview of some notation and terminology. Throughout the
paper, we use set theoretic notation for many concepts in logic. A \emph{clause}
is a set of literals that are part of an implicit disjunction. Similarly, a
\emph{formula} in CNF is a set of clauses that are part of an implicit
conjunction. We identify a \emph{model} with a set of variables that correspond
to the positive literals in the model (and all other variables are the negative
literals of the model). We can then define the \emph{cardinality} of a model as
the cardinality of this set. For example, let $\phi = (\neg a \lor b) \land a$
be a propositional formula over variables $a$ and $b$. Then an equivalent
set-theoretic representation of $\phi$ is $\{ \{ \neg a, b \}, \{ a \} \}$. Any
subset of $\{ a, b \}$ is an interpretation of $\phi$, e.g., $\{ a, b \}$ is a
model of $\phi$ (written $\{ a, b \} \models \phi$) of cardinality two, while
$\emptyset$ is an interpretation but not a model. We can now formally define
WMC.

\begin{definition}[WMC]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of \emph{indicator} variables, $X_P$ is the set of \emph{parameter}
  variables (with $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula
  in CNF over $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in
  X_I \cup X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of
  the instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

@@

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on initials of the
authors and the year of publication. Darwiche was the first to suggest the
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02} encoding that, in many ways, remains
the foundation behind most other encodings. He also introduced the distinction
between \emph{indicator} and \emph{parameter variables}; the former represent
variable-value pairs in the Bayesian network, while the latter are associated
with probabilities in the CPTs. The encoding \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05} is the only encoding that deviates from this
arrangement: for each variable in the Bayesian network, one indicator variable
acts simultaneously as a parameter variable. Chavira and Darwiche propose
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05} where they shift from WMC to
minimum-cardinality WMC because that allows the encoding to have fewer variables
and clauses. In particular, they propose a way to use the same parameter
variable to represent all probabilities in a CPT that are equal and keep only
clauses that `imply' parameter variables (i.e., omit clauses where a parameter
variable implies indicator variables). In their next encoding, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, the same authors optimise the aforementioned
implication clauses, choosing the smallest sufficient selection of indicator
variables. A decade later, Bart et al. present \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} that improves upon \texttt{cd06} in two ways.
First, they optimise the number of indicator variables used per Bayesian network
variable from a linear to a logarithmic amount. Second, they introduce a scaling
factor that can `absorb' one probability per Bayesian network variable. However,
for this work, we choose to disable the latter improvement since this scaling
factor is often small enough to be indistinguishable from zero without the use
of arbitrary precision arithmetic. Indeed, even a small Bayesian network with
seven mutually independent binary variables, 0.1 and 0.9 probabilities each, is
already big enough for the scaling factor to be exactly equal to zero (as
produced by the \texttt{bklm16}
encoder\footnote{\url{http://www.cril.univ-artois.fr/kc/bn2cnf.html}}).

@@

% TODO: briefly mention what a pseudo-Boolean function is
% TODO: mention that $2^x$ denotes the powerset of $X$.

In this work we propose a more expressive representation for WMC based on
pseudo-Boolean functions. Pseudo-Boolean functions, most commonly represented as
algebraic decision diagrams (ADDs) \cite{DBLP:journals/fmsd/BaharFGHMPS97}
(although a tensor-based approach has also been suggested
\cite{DBLP:journals/corr/abs-1908-04381,DBLP:conf/cp/DudekPV20}), have seen
extensive use in value iteration for Markov decision processes
\cite{DBLP:conf/uai/HoeySHB99}, both exact and approximate Bayesian network
inference \cite{DBLP:conf/ijcai/ChaviraD07,DBLP:conf/uai/GogateD11}, and
sum-product network \cite{DBLP:conf/uai/PoonD11} to Bayesian network conversion
\cite{DBLP:conf/icml/ZhaoMP15}. ADDs have been extended to compactly represent
additive and multiplicative structure \cite{DBLP:conf/ijcai/SannerM05},
sentences in first-order logic \cite{DBLP:journals/ai/SannerB09}, and continuous
variables \cite{DBLP:conf/uai/SannerDB11}, the last of which was also applied to
weighted model integration, i.e., the WMC extension for continuous variables
\cite{DBLP:conf/ijcai/BellePB15,DBLP:conf/ijcai/KolbMSBK18}.

Since two-valued pseudo-Boolean functions will be used extensively henceforth,
we introduce some notation. For any propositional formula $\phi$ over a set of
variables $X$ and $p, q \in \mathbb{R}$, let $[\phi]^p_q\colon 2^X \to
\mathbb{R}$ be a pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$. Next, we define some useful operations on
pseudo-Boolean functions.
% TODO: mention where the operations are from
% TODO: should I consider different domains?
% TODO: can I say that a domain automatically shrinks (w.r.t. x) if f|x=0 =
% f|x=1?

\begin{definition}[Operations] \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions, $x, y \in X$,
  $Y = \{y_i\}_{i=1}^n \subseteq X$, and $r \in \mathbb{R}$. Operations such as
  addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (f \cdot g)(Y)
    \coloneqq f(Y) \cdot g(Y).
  \]
  Note that this means that binary operations on pseudo-Boolean functions
  inherit properties such as associativity and commutativity. By regarding a
  real number as a constant pseudo-Boolean function, we can reuse the same
  definitions to define \emph{scalar} operations as
  \[
    (r+f)(Y) = r+f(Y), \quad \text{and} \quad (r \cdot f)(Y) = r \cdot f(Y).
  \]

  \emph{Restrictions} $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ of $f$ are
  defined as
  \[
    f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\}), \quad \text{and} \quad
    f|_{x=1}(Y) \coloneqq f(Y \cup \{x\})
  \]
  for all $Y \subseteq X$.

  \emph{Projection} $\exists_x$ is an endomorphism $\exists_x\colon
  \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined as
  \[
    \exists_xf \coloneqq f|_{x=1} + f|_{x=0}.
  \]
  Since projection is commutative (i.e., $\exists_x\exists_yf =
  \exists_y\exists_xf$) \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  we can define $\exists_Y\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ as
  $\exists_Y \coloneqq \exists_{y_1}\exists_{y_2}\dots\exists_{y_n}$. Throughout
  the paper, projection is assumed to have the lowest precedence (e.g.,
  $\exists_x fg = \exists_x (fg)$).
\end{definition}

Finally, below we list some properties of the operations on pseudo-Boolean
functions discussed in this section that can be conveniently represented using
our syntax. The proofs of all these properties follow directly from the
definitions.

@@

@@

\begin{definition}[PBP Instance] \label{def:new_wmc}
  A PBP instance is a tuple $(F, X, \omega)$, where $X$ is the set of variables,
  $F$ is a set of two-valued pseudo-Boolean functions $2^X \to \mathbb{R}$, and
  $\omega \in \mathbb{R}$ is the scaling factor. The \emph{answer} of the
  instance is $\omega \cdot \left(\exists_X\prod_{f \in F}f\right)(\emptyset)$.
\end{definition}

@@

In this section we describe an algorithm for transforming WMC instances to the
PBP format while removing all parameter variables. The algorithm works on four
out of the five Bayesian network encodings. There is no obvious way to adjust it
to work with \texttt{sbk05} because the roles of indicator and parameter (i.e.,
`chance') variables overlap \cite{DBLP:conf/aaai/SangBK05}. The algorithm is
based on several observations that will be made more precise in
\cref{sec:proof}. First, all weights except for $\{w(p) \mid p \in X_P\}$ are
redundant as they either duplicate an already-defined weight or are equal to
one. Second, each clause has at most one parameter variable. Third, if the
parameter variable is negated, we can ignore the clause (this idea first appears
in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}). Note that while
we formulate our algorithm as a sequel to the WMC encoding procedure primarily
because the implementations of Bayesian network WMC encodings are all
closed-source, as all transformations in the algorithm are local, it can be
efficiently incorporated into a WMC encoding algorithm with no slowdown.

@@

In this section we outline key properties that a (WMC or minimum-cardinality
WMC) encoding has to satisfy for \cref{alg:transformation} to output an
equivalent PBP instance. We divide the correctness proof into two theorems:
\cref{thm:correctness} for WMC encodings (i.e., \texttt{bklm16} and
\texttt{d02}) and \cref{thm:mccorrectness} for minimum-cardinality WMC encodings
(i.e., \texttt{cd05} and \texttt{cd06}). We begin by listing some properties of
pseudo-Boolean functions and establishing a canonical transformation from WMC
to PBP.

@@

\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that $\sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models
    \phi} \prod_{Y \models l} w(l)$ as required.
\end{proof}

@@

\Cref{cond:one} applies to all four WMC encodings under consideration and
ensures that only parameter variables carry nontrivial weights. \Cref{cond:d02}
(for \texttt{d02}), while spelled out precisely in the theorem, simply states
that each parameter variable is equivalent to a conjunction of indicator
literals. \Cref{cond:bklm16} is for encodings that have implications rather than
equivalences associated with parameter variables (which, in this case, is
\texttt{bklm16}). It ensures that each clause has at most one positive
parameter literal and no negative ones, and that at most one implication clause
per any parameter variable $p \in X_P$ can `force $p$ to be positive'.

@@

\section{Experimental Evaluation}
% TODO: what are the encodings compiled with?

>>>>

% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
% 1. Introduction. In one sentence, what’s the topic?
% 2. State the problem you tackle.
% 3. Summarize (in one sentence) why nobody else has adequately answered the research question yet.
% 4. Explain, in one sentence, how you tackled the research question.
% 5. In one sentence, how did you go about doing the research that follows from your big idea.
% 6. As a single sentence, what’s the key impact of your research?
\begin{abstract}
  Weighted model counting (WMC) is a powerful computational technique for a
  variety of problems, especially commonly used for probabilistic inference.
  However, the standard definition of WMC that puts weights on literals often
  necessitates WMC encodings to include additional variables and clauses just so
  each weight can be attached to a literal. This paper complements previous work
  by considering WMC instances in their full generality and using recent
  state-of-the-art WMC techniques based on pseudo-Boolean function manipulation,
  competitive with the more traditional WMC algorithms based on knowledge
  compilation and backtracking search. We present an algorithm that transforms
  WMC instances into a format based on pseudo-Boolean functions while
  eliminating around \SI{57}{\percent} of variables on average across various
  Bayesian network encodings. Moreover, we identify sufficient conditions for
  such a variable removal to be possible. Our experiments show significant
  improvement in WMC-based Bayesian network inference, outperforming the current
  state of the art.

  \keywords{Weighted model counting \and Probabilistic inference \and Bayesian
    networks.}
\end{abstract}

\section{Introduction}

@@

% NOTE: What's wrong with previous proposed solutions? How does mine differ?
One approach to this problem considers weighted clauses and probabilistic
semantics based on Markov networks \cite{DBLP:conf/uai/GogateD10}. However, with
a new representation comes the need to invent new encodings and inference
algorithms. Our work is similar in that it introduces a new representation for
computational problems but can reuse recent WMC algorithms based on
pseudo-Boolean function manipulation, namely, \textsf{ADDMC}
\cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC} \cite{DBLP:conf/cp/DudekPV20}.
Furthermore, we identify sufficient conditions for transforming a WMC instance
into our new format. As many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}, another way to avoid parameter variables
could be via direct compilation to a more convenient representation. However,
while direct compilation of Bayesian networks to SDDs has been investigated
\cite{DBLP:conf/ecsqaru/ChoiKD13}, SDDs are not expressive enough to avoid the
issue. Furthermore, to the best of the authors' knowledge, neither
compilation/inference technique has a publicly available implementation.

% NOTE: What are the key components of my approach and results? Also include any
% specific limitations. What about your project is new or different from
% existing solutions? List of refutable contributions (with forward references
% to the rest of the paper) What is the outcome of the project?
In this work, we introduce a way to transform WMC problems into a new format
based on pseudo-Boolean functions---\emph{pseudo-Boolean projection} (PBP). We
formally show that every WMC problem instance has a corresponding PBP instance
and identify conditions under which this transformation can remove parameter
variables. Four out of the five known WMC encodings for Bayesian networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}
can indeed be simplified in this manner. This transformation enables two
encodings that were previously incompatible with most WMC algorithms (due to
using a different definition of WMC
\cite{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}) to be run with
\textsf{ADDMC} and \textsf{DPMC} and results in a significant performance boost
for one other encoding, allowing it to surpass the current state of the art.
Finally, our theoretical contributions result in a convenient algebraic way of
reasoning about two-valued pseudo-Boolean functions and position WMC encodings
on common ground, identifying their key properties and assumptions.

@@

We begin with an overview of some notation and terminology. Throughout the
paper, we use set-theoretic notation for many concepts in logic. A \emph{clause}
is a set of literals that are part of an implicit disjunction. Similarly, a
\emph{formula} in CNF is a set of clauses that are part of an implicit
conjunction. We identify a \emph{model} with a set of variables that correspond
to the positive literals in the model (and all other variables are the negative
literals of the model). We can then define the \emph{cardinality} of a model as
the cardinality of this set. For example, let $\phi = (\neg a \lor b) \land a$
be a propositional formula over variables $a$ and $b$. Then an equivalent
set-theoretic representation of $\phi$ is $\{ \{ \neg a, b \}, \{ a \} \}$. Any
subset of $\{ a, b \}$ is an interpretation of $\phi$, e.g., $\{ a, b \}$ is a
model of $\phi$ (written $\{ a, b \} \models \phi$) of cardinality two, while
$\emptyset$ is an interpretation but not a model. We can now formally define
WMC.

\begin{definition}[WMC]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of \emph{indicator} variables, $X_P$ is the set of \emph{parameter}
  variables (with $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula
  in CNF over $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in
  X_I \cup X_P\} \to \mathbb{R}$ is the weight function. Its \emph{answer} is
  $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

@@

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on the initials of
the authors and the year of publication. Darwiche was the first to suggest the
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02} encoding that, in many ways, remains
the foundation behind most other encodings. He also introduced the distinction
between \emph{indicator} and \emph{parameter variables}; the former represent
variable-value pairs in the Bayesian network, while the latter are associated
with probabilities in the CPTs. The encoding \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05} is the only encoding that deviates from this
arrangement: for each variable in the Bayesian network, one indicator variable
acts simultaneously as a parameter variable. Chavira and Darwiche propose
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05} where they shift from WMC to
minimum-cardinality WMC because that allows the encoding to have fewer variables
and clauses. In particular, they propose a way to use the same parameter
variable to represent all probabilities in a CPT that are equal and keep only
clauses that `imply' parameter variables (i.e., omit clauses where a parameter
variable implies indicator variables). In their next encoding, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, the same authors optimise the aforementioned
implication clauses, choosing the smallest sufficient selection of indicator
variables. A decade later, Bart et al. present \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} that improves upon \texttt{cd06} in two ways.
First, they optimise the number of indicator variables used per Bayesian network
variable from a linear to a logarithmic amount. Second, they introduce a scaling
factor that can `absorb' one probability per Bayesian network variable. However,
for this work, we choose to disable the latter improvement since this scaling
factor is often small enough to be indistinguishable from zero without the use
of arbitrary precision arithmetic. Indeed, even a small Bayesian network with
seven mutually independent binary variables, 0.1 and 0.9 probabilities each, is
already big enough for the scaling factor to be exactly equal to zero (as
produced by the \texttt{bklm16}
encoder\footnote{\url{http://www.cril.univ-artois.fr/kc/bn2cnf.html}}).

@@

In this work, we propose a more expressive representation for WMC based on
pseudo-Boolean functions. A \emph{pseudo-Boolean function} is a function of the
form $\{ 0, 1 \}^n \to \mathbb{R}$ \cite{DBLP:journals/dam/BorosH02}.
Equivalently, let $X$ denote a set with $n$ elements (we will refer to them as
\emph{variables}), and $2^X$ denote its powerset. Then a pseudo-Boolean function
can have $2^X$ as its domain (then it is also known as a \emph{set function}).

Pseudo-Boolean functions, most commonly represented as
algebraic decision diagrams (ADDs) \cite{DBLP:journals/fmsd/BaharFGHMPS97}
(although a tensor-based approach has also been suggested
\cite{DBLP:journals/corr/abs-1908-04381,DBLP:conf/cp/DudekPV20}), have seen
extensive use in value iteration for Markov decision processes
\cite{DBLP:conf/uai/HoeySHB99}, both exact and approximate Bayesian network
inference \cite{DBLP:conf/ijcai/ChaviraD07,DBLP:conf/uai/GogateD11}, and
sum-product network \cite{DBLP:conf/uai/PoonD11} to Bayesian network conversion
\cite{DBLP:conf/icml/ZhaoMP15}. ADDs have been extended to compactly represent
additive and multiplicative structure \cite{DBLP:conf/ijcai/SannerM05},
sentences in first-order logic \cite{DBLP:journals/ai/SannerB09}, and continuous
variables \cite{DBLP:conf/uai/SannerDB11}, the last of which was also applied to
weighted model integration, i.e., the WMC extension for continuous variables
\cite{DBLP:conf/ijcai/BellePB15,DBLP:conf/ijcai/KolbMSBK18}.

Since two-valued pseudo-Boolean functions will be used extensively henceforth,
we introduce some new notation. For any propositional formula $\phi$ over $X$
and $p, q \in \mathbb{R}$, let $[\phi]^p_q\colon 2^X \to \mathbb{R}$ be the
pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$. Next, we define some useful operations on
pseudo-Boolean functions. The definitions of multiplication and projection are
equivalent to those in previous work
\cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20}.

\begin{definition}[Operations] \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions, $x, y \in X$,
  $Y = \{y_i\}_{i=1}^n \subseteq X$, and $r \in \mathbb{R}$. Operations such as
  addition and multiplication are defined pointwise, i.e., $(f+g)(Y) \coloneqq
  f(Y)+g(Y)$, and likewise for multiplication. Note that properties such as
  associativity and commutativity are inherited from $\mathbb{R}$. By regarding
  a real number as a constant pseudo-Boolean function, we can reuse the same
  definitions to define \emph{scalar} operations as $(r+f)(Y) = r+f(Y)$, and $(r
  \cdot f)(Y) = r \cdot f(Y)$.

  \emph{Restrictions} $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ of $f$ are
  defined as $f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\})$, and $f|_{x=1}(Y)
  \coloneqq f(Y \cup \{x\})$ for all $Y \subseteq X$.

  \emph{Projection} $\exists_x$ is an endomorphism $\exists_x\colon
  \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined as $\exists_xf \coloneqq
  f|_{x=1} + f|_{x=0}$. Since projection is commutative (i.e.,
  $\exists_x\exists_yf = \exists_y\exists_xf$)
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20}, we can define
  $\exists_Y\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ as $\exists_Y
  \coloneqq \exists_{y_1}\exists_{y_2}\dots\exists_{y_n}$. Throughout the paper,
  projection is assumed to have the lowest precedence (e.g., $\exists_x fg =
  \exists_x (fg)$).
\end{definition}

Below we list some properties of the operations on pseudo-Boolean functions
discussed in this section that can be conveniently represented using our syntax.
The proofs of all these properties follow directly from the definitions.

@@

\begin{remark}
  Note that our definitions of binary operations assumed equal domains. For
  convenience, we can assume domains to shrink whenever a function is
  independent of some of the variables (i.e., $f|_{x=0} = f|_{x=1}$) and expand
  for binary operations to make the domains of both functions equal. For
  instance, let $[x]_0^1,[\neg x]_0^1\colon 2^{\{x\}} \to \mathbb{R}$ and
  $[y]_0^1\colon 2^{\{y\}} \to \mathbb{R}$ be pseudo-Boolean functions. Then
  $[x]_0^1 \cdot [\neg x]_0^1$ has $2^\emptyset$ as its domain. To multiply
  $[x]_0^1$ and $[y]_0^1$, we expand $[x]_0^1$ into $\left([x]_0^1\right)'\colon
  2^{\{x, y\}} \to \mathbb{R}$ which is defined as $\left([x]_0^1\right)'(Z)
  \coloneqq [x]_0^1(Z \cap \{ x \})$ for all $Z \subseteq \{ x, y \}$ (and
  equivalently for $[y]_0^1$).
\end{remark}

@@

\begin{definition}[PBP Instance] \label{def:new_wmc}
  A PBP instance is a tuple $(F, X, \omega)$, where $X$ is the set of variables,
  $F$ is a set of two-valued pseudo-Boolean functions $2^X \to \mathbb{R}$, and
  $\omega \in \mathbb{R}$ is the scaling factor. Its \emph{answer} is $\omega
  \cdot \left(\exists_X\prod_{f \in F}f\right)(\emptyset)$.
\end{definition}

@@

In this section, we describe an algorithm for transforming WMC instances to the
PBP format while removing all parameter variables. The algorithm works on four
out of the five Bayesian network encodings. There is no obvious way to adjust it
to work with \texttt{sbk05} because the roles of indicator and parameter (i.e.,
`chance') variables overlap \cite{DBLP:conf/aaai/SangBK05}. The algorithm is
based on several observations that will be made more precise in
\cref{sec:proof}. First, all weights except for $\{w(p) \mid p \in X_P\}$ are
redundant as they either duplicate an already-defined weight or are equal to
one. Second, each clause has at most one parameter variable. Third, if the
parameter variable is negated, we can ignore the clause (this idea first appears
in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}). Note that while
we formulate our algorithm as a sequel to the WMC encoding procedure primarily
because the implementations of Bayesian network WMC encodings are all
closed-source, as all transformations in the algorithm are local, it can be
efficiently incorporated into a WMC encoding algorithm with no slowdown.

@@

In this section, we outline key properties that a (WMC or minimum-cardinality
WMC) encoding has to satisfy for \cref{alg:transformation} to output an
equivalent PBP instance. We divide the correctness proof into two theorems:
\cref{thm:correctness} for WMC encodings (i.e., \texttt{bklm16} and
\texttt{d02}) and \cref{thm:mccorrectness} for minimum-cardinality WMC encodings
(i.e., \texttt{cd05} and \texttt{cd06}). We begin by listing some properties of
pseudo-Boolean functions and establishing a canonical transformation from WMC
to PBP.

@@

\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the WMC answer is
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that $\sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models
    \phi} \prod_{Y \models l} w(l)$ as required.
\end{proof}

@@

\Cref{cond:one} applies to all four WMC encodings under consideration and
ensures that only parameter variables carry nontrivial weights. \Cref{cond:d02}
(for \texttt{d02}) simply states that each parameter variable is equivalent to a
conjunction of indicator literals. \Cref{cond:bklm16} is for encodings that have
implications rather than equivalences associated with parameter variables
(which, in this case, is \texttt{bklm16}). It ensures that each clause has at
most one positive parameter literal and no negative ones, and that at most one
implication clause per any parameter variable $p \in X_P$ can `force $p$ to be
positive'.

@@

\section{Experimental Evaluation}

We run a set of experiments, comparing all five original Bayesian network
encodings (\texttt{bklm16}, \texttt{cd05}, \texttt{cd06}, \texttt{d02}
\texttt{sbk05}) as well as the first four with \cref{alg:transformation} applied
afterwards.\footnote{Recall that \texttt{cd05} and \texttt{cd06} are
  incompatible with \textsf{DPMC}.} For each encoding \texttt{e}, we write
\texttt{e++} to denote the combination of encoding a Bayesian network as a WMC
instance using \texttt{e} and transforming it into a PBP instance using
\cref{alg:transformation}. Along with
\textsf{DPMC}\footnote{\url{https://github.com/vardigroup/DPMC}}, we also
include WMC algorithms used in the papers that introduce each encoding:
\textsf{Ace} for \texttt{cd05}, \texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\cite{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and
\textsf{c2d}\footnote{\url{http://reasoning.cs.ucla.edu/c2d/}}
\cite{DBLP:conf/ecai/Darwiche04} with
\textsf{query-dnnf}\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
for \texttt{bklm16}. \textsf{Ace} is also used to encode Bayesian networks into
WMC instances for all encodings except for \texttt{bklm16} which uses another
encoder mentioned previously. We focus on the following questions:
\begin{itemize}
\item Can parameter variable elimination improve inference speed?
\item How does DPMC combined with encodings without (and with) parameter
  variables compare with other WMC algorithms and other encodings?
\item Which instances is our approach particularly successful on (compared to
  other algorithms and encodings and to the same encoding before our
  transformation)?
\item What proportion of variables is typically eliminated?
\item Do some encodings benefit from this transformation more than others?
\end{itemize}
